---
title: "SPEW Development Process"
author: "Lee Richardson"
date: "January 2, 2017"
output: html_document
---

This document details the SPEW development process. It is written for people who are adding code to the SPEW package. After reading, contributors should understand the organization of SPEW software, and how to effectively contribute. The goal is for all SPEW software to be consistent, robust, modular, and easily extendable.

The SPEW development process is by no means optimal. We are consistently looking for new approaches for a more efficient process. However, we need to ensure that changes to the process do not interfere with consistency and correctness, since SPEW ecosystems and software have already been publically released. 

The SPEW process is conservative, with many checks and balances for ensuring that code which is added to the current release has passed several tests. 

# SPEW Overview and Organization
SPEW software serves two primary goals:

1. Generating synthetic ecosystems for release on Olympus 
2. Providing general-purpose tools for custom synthetic ecosystems 

To meet goal 1, a separate directory **olympus/** contains all of the code corresponding to ecosystem generation on olympus. The Olympus directory has the code for running SPEW on olympus, generating reports, collecting data, etc. In addition, the spew geographic hierarchy is here. More information on Olympus is provided below. 

To meet goal 2, SPEW software is organized into an R package. The primary reference SPEW uses is the book **R Packages**, by Hadley Wickham. The book is available online at: <http://r-pkgs.had.co.nz/>. The majority of this document details the specifics of package development, tailored to SPEW. In particular, if you do not understand a component of the SPEW package, this book is often the first place to look. 

# R Code 
Since SPEW is an R-package, a big part of the development process is writing R-code. Follow R-package conventions, all R-code is located in the *R/* directory. Within the *R/* directory, there are *.R* files, with functions implementing the SPEW framework. An important point is that all code in **R/** are functions, rather than scripts. This is because (almost) everything is R is a function call, so the primary code in packages are functions. Each *.R* file contains a group of related functions. For example, *R/ipf.R* has all functions related to IPF sampling, and *R/read.R* has all functions for reading input data. 

## Consistency 
A goal of SPEW software moving forward is consistent syntax. To achieve consistent syntax, we follow the Style guide described in Hadley Wickham's book, [Advanced R](http://adv-r.had.co.nz/Style.html). Admittedly, we are not fully compliant with the style guide right now. But, hopefully we can keep iterating and fixing the SPEW code. The point of the Style guide isn't to be pedantic and dogmatic, but rather for ensuring that code is consistent, regardless of the contributor. Often, it is difficult to read or understand someone else's code if the style/syntax is different. This is highly inefficient, since a lot work goes into parsing, testing, and understanding the code. Following a consistent syntax alleviates this issue.

Here is an example function from the *R/ipf.R* file. This is the main IPF-wrapper function, which calls smaller IPF-related functions for carrying out the complete IPF sampling procedure. 

```{r, eval = FALSE}
sample_ipf <- function(n_house, pums_h, pums_p, marginals, alpha = 0, k = .001, 
                       puma_id = NULL, place_id = NULL, do_subset_pums = TRUE) {
                                        # Step 1: Align PUMS with Marginals

    if(do_subset_pums){
        pums <- subset_pums(pums_h = pums_h, pums_p = pums_p, marginals = marginals, puma_id = puma_id)
    } else {
        pums <- pums_h
    }
    pums <- align_pums(pums, marginals)


  # Step 2: Fill in the contingency table
    table <- fill_cont_table(pums = pums, marginals = marginals, place_id = place_id, n_house = n_house)

                                        # Write out the contingency table HERE.
  
  # Step 3: Sample with contingency table weights 
  households <- sample_with_cont(pums = pums, table = table, alpha = alpha, 
                                 k = k, marginals = marginals)
  
  return(households)
}
```

There are three steps involved in the procedure:

1. Aligning the PUMS microdata with the Marginal tables 
2. Estimating the contingency table 
3. Sampling using the contingency table weights 

Over-time, the function has been modified in order to meet various deadlines. To see this, we can look at the history of changes to this particular file: <https://github.com/leerichardson/spew/commits/master/R/ipf.R>. This section walks through some of the issues with this function, and re-factors the function back into style-guide compliance. Of course, all of this must be done while retaining the correctness of the function. Ultimately, we want all SPEW code following consistent syntax. While this will take effort, it will go a long way towards making SPEW readable and reliable moving forward. 

Immediately, we can see problems with the function. First, some of the comments have been moved to the far-right corner of the line. In addition, step 2 of the procesure and the corresponding comment aren't aligned. Finally, the syntax surrounding the if statement doesn't follow the spacing section of the Style guide. Generally speaking, spacing within a function should be consistent, and every line should start with the same indent. In addition, coding syntax should follow the style guide. Fixing these issues:

```{r, eval = FALSE}
sample_ipf <- function(n_house, pums_h, pums_p, marginals, alpha = 0, k = .001, 
                       puma_id = NULL, place_id = NULL, do_subset_pums = TRUE) {
  # Step 1: Align PUMS with Marginals
  if (do_subset_pums) {
      pums <- subset_pums(pums_h = pums_h, pums_p = pums_p, marginals = marginals, puma_id = puma_id)
  } else {
      pums <- pums_h
  }
  pums <- align_pums(pums, marginals)

  # Step 2: Fill in the contingency table
  table <- fill_cont_table(pums = pums, marginals = marginals, place_id = place_id, n_house = n_house)
  # Write out the contingency table HERE.
  
  # Step 3: Sample with contingency table weights 
  households <- sample_with_cont(pums = pums, table = table, alpha = alpha, 
                                 k = k, marginals = marginals)
  
  return(households)
}
```

Which already looks much better. In particular, the structure of the function is now much more clear, and three step procedure is obvious to anyone reading it. 

## If statements 
The `sample_ipf` function provides room to explore another *important* point about the R-code, if-statements. Generally speaking, we want to minimize the amount of ad-hoc, if-statements in the SPEW code. However, in many situations if-statements are the only method capable of solving the problem. So, we face a tradeoff. The main concept is that even if an if-statement is necessary, its purpose should be well documented, and it should be obvious to anyone reading the code why the if-statement should be there. 
  
Upon reviewing the history, we can see the if-statement in `sample_ipf` was added during the following commit message:

> "made changes to IPF to use on a custom population. There was also an issue with a single variable IPF that I fixed."

The main problem is that no-where, either in the code or the commit message, is the issue described. So, it is unclear _why_ this particular if-statement fixes the problem. Clearly, this was added in a rush in order to meet a deadline. While this is ok, too much reliance of this anxiety-ridden approach will slowly deteriorate SPEW. For SPEW to be maintainable, a better balance must be struck between _getting it done_ and maintaining the quality of the software. 

Finally, since the if-statement clearly corresponds to the smaller `subset_pums` function, it is more appropriate to implement there. This way, the top-level wrapper function is still readable, and the if-statement occurs at the level where it actually takes place. With these changes, the function is restored!

```{r, eval = FALSE}
sample_ipf <- function(n_house, pums_h, pums_p, marginals, alpha = 0, k = .001, 
                       puma_id = NULL, place_id = NULL, do_subset_pums = TRUE) {
  # Step 1: Align PUMS with Marginals
  pums <- subset_pums(pums_h = pums_h, pums_p = pums_p, marginals = marginals, puma_id = puma_id)
  pums <- align_pums(pums, marginals)

  # Step 2: Fill in the contingency table
  table <- fill_cont_table(pums = pums, marginals = marginals, place_id = place_id, n_house = n_house)
  # Write out the contingency table HERE.
  
  # Step 3: Sample with contingency table weights 
  households <- sample_with_cont(pums = pums, table = table, alpha = alpha, 
                                 k = k, marginals = marginals)
  
  return(households)
}
```


In sum, a strategy for if statements 

1. Avoid ad-hoc if statements if possible
2. If the if-statement is necessary, include it at the appropriate level (not the top-level wrapper function)
3. Clearly describe *why* the if-statement is necessary

## Using defaults 
- Use defaults to add new features 

## CRAN Specific Conventions 
- If using another function, use package::function 
- If using a "Suggests" package, do requireNamespace


# Unit Testing 



# Documentation 
SPEW uses three different layers of documentation. First, the R-code if documented accoring to the `roxygen2` package. 




## Functions 



## New Data for SPEW ecosystem releases 



## New Methods for SPEW ecosystem release 



# Automated Checks 
In the R-world, a package truly _makes it_ when it is avaialable on CRAN. CRAN stands for the Comprehensive R Archive Network,  


## Integration with Travis 



# Adding Major Features 



## Branches



## Code Reviews 



# Olympus 
- .Rbuildignore `(devtools::use_build_ignore("olympus"))`

