---
title: "The SPEW Process"
author: "Lee Richardson"
date: "January 2, 2017"
output: html_document
---

This document details the SPEW development process. It is written for people adding code to the SPEW package. After reading, contributors should understand the organization of SPEW and how to contribute. The goal is for all SPEW software to be consistent, robust, and modular. 

The process is by no means optimal, and we are interested in improvements. However, we need to guard against introducing error, since SPEW ecosystems are already publically available. Therefore, the SPEW process contains many checks and balances for ensuring new code is as correct as possible.

This document is organized as follows. We first describe SPEW's organization, followed by the R-code structures, and conclude with the documentation system. 

# SPEW Overview 
SPEW serves two purposes:

1. Generating and releasing synthetic ecosystems 
2. Providing a general-purpose synthetic ecosystem generator 

To meet goal 1, we generate and host our synthetic ecosystems on the Olympus computing cluster, hosted by the Pittsburgh Supercomputing Center (PSC). To meet goal 2, we organize SPEW into an R-package, avaialble for users with  custom data. The next section details code for both code for Olympus and the R-package.

# SPEW Organization 
All SPEW code is on Github at <https://github.com/leerichardson/spew>. Here, the **olympus/** directory contains all code used for generating and hosting our ecosystems. The **doc/** directory contains our formal documentation, and everything else corresponds to code for the R-package. 

## Olympus 
The **olympus/** directory contains code for running SPEW on Olympus, generating reports, collecting and formatting data, analyzing logfiles, and generating the SPEW geographic hierarchy. All of our files are located on Olympus at: 

  **/mnt/beegfs1/data/shared_group_data/syneco**.

Here, directories with names _spew_NUMBER_ are releases of SPEW ecosystems data. The directory **olympus/** synchronizes with the **olympus/** directory on Github. Lee uses the file **olympus/move_file.bash** for moving code back and forth, which is easily adapted to other computers. Files we no longer use are in the **old/** directory.

Our final ecosystems are available online at: <http://data.olympus.psc.edu/syneco/>. This website simply mirrors the Olympus directory:

  **/mnt/lustre0/machines/data.olympus.psc.edu/srv/apache/data/syneco**

So, anything we want to appear online, we can just put here. We currently post finalized versions of our ecosystems using "symlinks" between this directory and the directory all of our files are in, to avoid copying. 

We now briefly explain each subdirectory inside **olympus/**. 

* **call_spew/**: All code for generating synthetic ecosystems on Olympus using the *spew* package
* **data/**: All code for downloading and pre-procesing data. Still needs to be moved from **old/spew_olympus**  
* **logfiles/**: Code for extracting and analyzing logfiles from SPEW output. Used to generate timings in the paper.
* **reports/**: Code for generating diagnostic reports
* **spew_hierarchy/**: Code for generating our geographic directories 
* **misc/**: Everything else, including for custom ecosystems 

## R-Package 
The primary reference for our *spew* R-package is  **R Packages**, available online at: <http://r-pkgs.had.co.nz/>.
Most of our structure follows the structure from this book. For new contributors, the two most important parts are the **R/** folder, where all R-code lives, and **tests/**  for verifying accuracy of our R-code. The primary concept  to remember is that all code should have a corresponding test, and all tests should pass before adding code. The **data/** and **data-raw/** folders have example data, which can be used for testing new code.

# R-Code 
We (ideally) follow the style guide available from R-packages at: <http://r-pkgs.had.co.nz/r.html>. Unfortunately, previous spew code was hastily written, so our package is not entirely consistent. Please strive to follow these conventions when adding code, since this makes things much simpler to understand. Additionally, please make sure code is adequately commented, meaning that anyone reading it should be able to understand why and how things are what they are. If you are a new contributor, we are available to review and comment on new additions. 

An example of re-factoring some existing code is contained in the appendix. 

## Functions 
The **R/** directory has files for various groupings of functions. It is important to stress that all R-code in these files should be functions, not scripts. Ideally, the functions should work as generally as possible, and shouldn't have specific functionality tailored to Olympus file-paths.

The main function is in the **spew.R** file, containing the **call_spew** function we use on olympus, and the **spew** function used more generally. The **spew** function essentially loops rows of the population table, generating a synthetic population for each row. We also include different wrappers for running this in parallel. 

## Unit Tests
Key to keeping our functions correct are unit-tests. Unit testing is covered in detail here  <http://r-pkgs.had.co.nz/tests.html>. All functions added to *spew* should have a corresponding tests, and all test must pass before code is added. If it is the end of the day, and the code still doesn't work, it is better to wait than put it online without testing. 

### Integration with Travis CI 
It would be great to integrate SPEW with the Travis Continuous Integration system. This way, every time code is added, all automatic checks required by CRAN (and our tests) are automatically ran, and we are notified if the new code breaks any of our existing tests. We hope to incorporate this soon.

#### Minor Details  
- If using a function from another package, use package::function 
- If using a "Suggests" package, use a requireNamespace condittional before calling 

# Documentation 
SPEW has two layers of documentation: for R-functions, and a more formal documentation for each public release. Both are detailed next. 

## R-Functions 
Each R-function should be documented, following  <http://r-pkgs.had.co.nz/man.html>. The documentation should be as concise as possible. Additionally, remember to move to a new-line when writing documentation, and try to keep less than 100 characters per line. 

## SPEW Ecosystems 
Each release of SPEW ecosystems codes with a corresponding *.pdf* documentation, located at **doc/ecosystems/VERSION**. While the template and wording remains similar, we need to revise this carefully for each release, and make sure all relevant details are contained here. The following are components we need in each set of *.pdf* documentation.

### New Data Sources  
All data-sources should used in this release must be documented here. If you are adding new data, you must document it here. Recall that all code for downloading and pre-processing of data should be made available at: **olympus/data**. 

### New Methodology 
All methodology for each component *spew* must be detailed here. If you are adding new methodology, it must be formally documented here. 

### Clarifications for this release 
Details of the release, including anything done in an ad-hoc manner, should be recorded. For example, we might need to record that "For canada, we adapted the version of spatial sampling because there was no lake data". 

### Description of file-formatting
We have a complicated directory for each ecosystem, which should be layed out here. 

# Appendix 
Extraneous details on the SPEW process are given here. 

## Re-factoring R Code 
This section revises an existing *spew* R-function for compliance with the style guide. We hope this provides inspiration for how to get the code looking good, and eventually revise all code for consistency.

Here is an example function from the *R/ipf.R* file. This is the IPF-wrapper function, sequentially calling functions implementing smaller pieces of the IPF method. There are three steps involved in the procedure:

1. Aligning the PUMS microdata with the Marginal tables 
2. Estimating the contingency table 
3. Sampling using the contingency table weights 

```{r, eval = FALSE}
sample_ipf <- function(n_house, pums_h, pums_p, marginals, alpha = 0, k = .001, 
                       puma_id = NULL, place_id = NULL, do_subset_pums = TRUE) {
                                        # Step 1: Align PUMS with Marginals

    if(do_subset_pums){
        pums <- subset_pums(pums_h = pums_h, pums_p = pums_p, marginals = marginals, puma_id = puma_id)
    } else {
        pums <- pums_h
    }
    pums <- align_pums(pums, marginals)


  # Step 2: Fill in the contingency table
    table <- fill_cont_table(pums = pums, marginals = marginals, place_id = place_id, n_house = n_house)

                                        # Write out the contingency table HERE.
  
  # Step 3: Sample with contingency table weights 
  households <- sample_with_cont(pums = pums, table = table, alpha = alpha, 
                                 k = k, marginals = marginals)
  
  return(households)
}
```

We immediately spot problems with the function: 

1. Some comments have been moved to the far-right corner of the line. 
2. Step 2 of the procedure and the corresponding comment are not aligned. 
3. The syntax surrounding the if statement doesn't follow the spacing section of the Style guide. 

Fixing these issues:

```{r, eval = FALSE}
sample_ipf <- function(n_house, pums_h, pums_p, marginals, alpha = 0, k = .001, 
                       puma_id = NULL, place_id = NULL, do_subset_pums = TRUE) {
  # Step 1: Align PUMS with Marginals
  pums <- subset_pums(pums_h = pums_h, pums_p = pums_p, marginals = marginals, puma_id = puma_id)
  pums <- align_pums(pums, marginals)

  # Step 2: Fill in the contingency table
  table <- fill_cont_table(pums = pums, marginals = marginals, place_id = place_id, n_house = n_house)
  # Write out the contingency table HERE.
  
  # Step 3: Sample with contingency table weights 
  households <- sample_with_cont(pums = pums, table = table, alpha = alpha, 
                                 k = k, marginals = marginals)
  
  return(households)
}
```

We can extract various principles out of this revision

### Consistency
We want *spew* code following consistent syntax. Not to be pedantic or dogmatic, but rather for keeping consistency across contirbutors. It is ofter difficult to understand someone else's code following a different coding style or syntax. This is highly inefficient, since a lot work goes into parsing, testing, and understanding the code. Following a consistent syntax alleviates this issue.

### If-statements 
Generally speaking, we want to minimize ad-hoc if-statements in the SPEW code. However, in many situations if-statements are the only method for solving the problem. When if-statements are required, we really want it to be well documented, so anyone reading it knows why it should be there.

A strategy for if-statements:

1. Avoid ad-hoc if statements if possible
2. If the if-statement is necessary, include it at the appropriate level (not the top-level wrapper function)
3. Clearly describe *why*, if the if-statement is nececssary the if-statement is necessary

### Using defaults to add new-features 
When adding to *spew*, make the existing method a default. Defaults allow us to update the code in one place, avoiding having to update a million smaller locations anytime something changes. 

